Mainstreaming Race Science

### 2023-07-19 paper/paper.Rmd
mostly filling in supplement refs

---
title: "Mainstreaming Race Science"
authors:
  - name: "Daniel J. Hicks"
    affiliation: 1
    corresponding_author: true
  - name: "Emilio Lobato"
    affiliation: 1
address:
  - code: 1
    address: "University of California, Merced"
corresponding_author: "Corresponding author: Daniel J. Hicks, dhicks4@ucmerced.edu."
abstract: |
  *[structured]*
acknowledgements: |
  *[strutured]* Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  EL's work on this project was supported by UC Merced.  DH's work on this project was not supported by any specific funding. 
bibliography: "references.bib"
output: 
  # https://github.com/rstudio/rticles/pull/486
  # skeleton: <https://github.com/christopherkenny/rticles/tree/master/inst/rmarkdown/templates/science>
  # rticles::science_article:
  #   keep_tex: true
  #    move_figures: false
  #    draft: false
  #    number_sections: false
  bookdown::pdf_book:
    base_format: rticles::science_article
    keep_tex: true
    move_figures: false
    draft: false
    number_sections: false
---

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results *[date coverage]* (\#fig:wos)](img/wos_results.png){width="4.76in" height="2.6in"}

@@

Our findings indicate that *Mankind Quarterly* (MQ) was less important as an echo chamber for late 20th century race science than mainstream psychology journals, especially *Intelligence* (I) and *Personality and Individual Differences* (PID).  Indeed, MQ published a very different kind of race science from the race-and-intelligence research published in I and PID, even after *[Lynn joined the MQ editorial board]*.

*[for discussion: the possibility for white supremacist appropriation, ie, race science done by scientists who are not themselves scientific racists: CarlsonQuantifyingContextualizingImpact2020; probably some reviews of Paige Harden; O'Connor and Weatherall]*

We stress that we are not identifying any particular scientist as a racist, scientific racist, or race scientist based on the topic model results.  Scientific research can be utilized to promote scientific racism — and thus count as race science — independently of the views and attitudes of the scientist who conducted that research.\cite{TaberyWhyStudyingGenetics2015, GillbornSoftlySoftlyGenetics2016, CarlsonQuantifyingContextualizingImpact2020, HennWhyDNANo2021}  But this does not mean that scientists are not responsible for the downstream social effects of their 
research, even when these effects involve deliberate misinterpretation.*[Douglas, Kitcher, Kourany, Havstad; Block and Dworkin?]*  Like other citizens, scientists are responsible for mitigating the reasonably foreseen harms that result from their actions.  And we now have decades of empirical evidence that research on the genetics of intelligence — even research that does not itself touch on race — can and will be used to promote scientific racism \cite{MeyerWrestlingSocialBehavioral2023}.  Specifically, while our findings suggest that behavior genetics has attempted to distance itself from scientific racism, behavior genetics research — and research on human genetics more generally — may be especially susceptible to appropriation by scientific racists \cite{CarlsonQuantifyingContextualizingImpact2020, OgbunugaforDNABasketballBirthday2022}.  A colorblind attitude of "we don't talk about race anymore" *[cite Bonilla-Silva?]* is likely insufficient to prevent particular research findings being utilized as race science.

@@

\appendix
# Materials and methods

*[probably some introduction]*

@@

Corpus assembly is summarized in figure *[xref]*.  The corpus was assembled in two parts, one for *Mankind Quarterly* and the other for mainstream journals.

*[figure]*

@@

An electronic archive of *Mankind Quarterly*, from the first issue in *[1969]* through 2004, is available for free on the open web from *The Unz Review*, which has been characterized as a white nationalist website by the Southern Poverty Law Center *[<https://www.splcenter.org/hatewatch/2021/01/19/meet-white-nationalist-organizer-who-spewed-hate-against-lawmakers>]*.  Web scraping packages and techniques *[which?]* were used to retrieve every PDF available in this archive.  These PDFs were processed using the open-source OCR (optical character recognition) software `tesseract` and the R API `rtesseract` *[cites]*.  This stage of corpus assembly was conducted in Fall 2021.

@@

To identify suitable mainstream journals for inclusion in the corpus, we first identified academic researchers who had been funded by the Pioneer Fund.  Pioneer is an American non-profit organization founded in 1937 to fund eugenics research *[cite]*.  After the *Brown v. Board of Education* ruling in 1954, Pioneer funded various segregationist efforts across the United States.  *[Shockley and Jensen]*.  We reviewed a critical profile of Pioneer *[cite]* as well as an archived page from the organization's own web site *[cite]*, which together listed 16 researchers who had received Pioneer funds.

We then used the author search tool in Clarivate's Web of Science platform *[url]*, retrieving publication lists for 14 researchers.  These searches were conducted between 2021-09-24 and 2021-10-05 by DJH.  After parsing these results, we counted how many of the 14 researchers had published in each journal.  13 journals had published 6 or more of the 14 Pioneer-funded researchers.  We excluded *Mankind Quarterly* (as already included) as well as *Science* and *Nature* (as too general) from further consideration in this side of the corpus.  3 journals published by the American Psychological Association (APA; *American Psychologist*, *Contemporary Psychology*, and *Journal of Educational Psychology*) had to be excluded due to confusion over who could give us permission to use the archives for a text mining project, with both APA and ProQuest asserting that we needed to get permission from the other entity.  *European Journal of Personality*, published by SAGE, also had to be excluded because our institutional access only went back to 1999.

The remaining 5 journals are all published by major academic publishers — Elsevier, Springer, or Cambridge University Press — and each item in the entire run of each journal has been assigned a DOI (digital object identifier) for archival purposes.  We used the Crossref API and `rcrossref` R interface to this API *[cite]* to retrieve metadata for each item published from 1960-2010 in each of these 6 journals.  These metadata included item-level license information — confirming that the text of each item could be used for text mining projects — and a URL to an electronic version of the item.  These URLs were used to retrieve an HTML or PDF version of each item, except for *Personality and Individual Differences*.  This journal has published a relatively large number of non-article documents, such as book reviews and commentaries, which are not available at the URL included in the Crossref metadata.  (This is unfortunate, as it was not difficult to find highly relevant documents that we could not automatically retrieve and therefore excluded from our corpus *[review]*.)  Instead we used Elsevier's ScienceDirect API *[url]* to search and retrieve all items from *Personality and Individual Differences* that are available.  PDFs were run through OCR and text extraction as necessary, as with the *Mankind Quarterly* documents.

All together, *[corpus summary, total and by journal; table]*

@@

After document retrieval and text extraction, we pre-processed the text using the `spaCy` NLP (natural language processing) Python library and the R API `spacyr` *[version, cites]*.  Specifically, we applied regular expressions to remove header/footer copyright notices and hyphenation, used spaCy to annotate and extract noun phrases (eg, "the intelligence test items"), and then cleaned and standardized these phrases (eg, removing the/an/a, coverting all text to lowercase, and replacing all whitespace with underscores: "intelligence_test_items").  We then counted the occurrence of each noun phrase in the document.  The aggregated "document-term matrix" was stored in Parquet format *[cite]* for performance reasons, and written and read using the `arrow` package for R *[cite]*.  *[availability]*

NLP-extracted noun phrases offer a number of advantages over the more traditional unigram ("single-word") tokens.  First, noun phrase extraction removes many standard stopwords (articles, common verbs) without relying on a fixed, a priori stopword list.  Phrases can be more informative than single terms, for example, distinguishing "intelligence test" from "hypothesis test."  But simple n-gram extraction will include numerous phrases that are not especially meaningful.  Consider the sentence "Since our first analyses of feeding patterns in rats, we had been using a criterion of 40 minutes (Le Magnen & Tallon 1963; 1966)" *[10.1017__^__s0140525x0000042x]*.  Bigrams such as "since our" and "criterion 40" will likely be discarded in vocabulary selection, but significantly increase the computational cost of vocabulary selection.  Noun phrase extraction is therefore more efficient.

After noun phrase extraction, the corpus comprised *[tokens]* of *[distinct noun phrases]*.

@@

We took an information-theoretic approach to vocabulary selection.  Consider a game in which I draw a document from the corpus, then a single token from that document.  I tell you the term, and you have to guess which document I picked.  Intuitively, highly informative terms (in this project, noun phrase types) are distinctive, allowing you to dramatically narrow down the list of potential documents.  This "informativeness" of a term can be quantified as the KL (Kullback-Leibler) divergence from a "baseline" distribution of documents to the distribution conditional on the term.  Because the most informative terms tend to be typos and OCR errors — these are unique to a single document — we multiply the KL divergence by the logarithm of the total number of occurrences of the term across the entire corpus.  We refer to the resulting measure as $log(n) \Delta H$ or simply `ndH`.

*[Hicks 2021?]* used the `ndH` approach in a topic modeling analysis, where the baseline distribution was the uniform distribution across documents.  In the current project, we found that this approach heavily favored recurrent noun phrases in the longest documents.  Many of the documents published in *Psychological Reports* are extremely short, 1-2 reports of a single study; while many of the documents published in *Brain and Behavioral Sciences* are book-length collections that include a long review article and sometimes dozens of commentaries.  Very generic noun phrases that happen to appear in the latter can occur orders of magnitude more often than highly distinctive phrases in the former, and so the $log(n)$ factor overwhelms the $\Delta H$ factor.

@@

A common rule of thumb in topic modeling is that the vocabulary should have about 10 times as many distinct terms as the number of documents in the corpus *[cite]*.  However, we had some concerns with computational demands here:  the resulting document-term matrix would have roughly $10 \times 33,000^2$ or 10.9 billion entries; with 10% density this would require on the order of 4 GB of memory just for a single copy of the matrix; and most of the analysis was to be conducted on the authors' laptops.  We therefore chose to work with three smaller vocabularies, $5 \times$, $1 \times$ and $\frac{1/5} \times$ the number of documents.  We refer to these as the "large," "medium," and "small" vocabulary, respectively, and they include *[counts]* distinct phrases.  We compare findings across vocabularies as a robustness check.

>>>>

---
title: "Mainstreaming Race Science"
authors:
  - name: "Daniel J. Hicks"
    affiliation: 1
    corresponding_author: true
  - name: "Emilio Lobato"
    affiliation: 1
address:
  - code: 1
    address: "University of California, Merced"
corresponding_author: "Corresponding author: Daniel J. Hicks, dhicks4@ucmerced.edu."
abstract: |
  *[structured]*
acknowledgements: |
  *[strutured]* Thanks to Emily Merchant and John Jackson for some initial discussions that helped clarify the scope of the project and identify some key resources related to the Pioneer Fund and *Mankind Quarterly*.  Thanks to Anthony Sainez for help retrieving the *Mankind Quarterly* articles.  Thanks to Derek Devnich and James Dooley for their work attempting to secure the APA-published articles.  EL's work on this project was supported by UC Merced.  DH's work on this project was not supported by any specific funding. 
bibliography: "references.bib"
output: 
  # https://github.com/rstudio/rticles/pull/486
  # skeleton: <https://github.com/christopherkenny/rticles/tree/master/inst/rmarkdown/templates/science>
  bookdown::pdf_book:
    base_format: rticles::science_article
    keep_tex: true
    move_figures: false
    draft: true
    number_sections: false
---

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results. (\#fig:wos)](img/wos_results.png){width="4.76in" height="2.6in"}

@@

Using bibliometric and text-mining methods, this study finds that Pioneer-funded researchers published heavily in certain mainstream psychology journals, rather than *Mankind Quarterly*; and that a distinct topic of race science discourse can be identified in these same journals across the period 1960-2010.  These findings indicate that *Mankind Quarterly* (MQ) was less important as an echo chamber for late 20th century race science than mainstream psychology journals, especially *Intelligence* (I) and *Personality and Individual Differences* (PID).  Indeed, MQ published a very different kind of race science from the race-and-intelligence research published in I and PID.

We stress that we are not identifying any particular scientist as a racist, scientific racist, or race scientist based on the topic model results.  Scientific research can be utilized to promote scientific racism — and thus count as race science — independently of the views and attitudes of the scientist who conducted that research.\cite{TaberyWhyStudyingGenetics2015, GillbornSoftlySoftlyGenetics2016, CarlsonQuantifyingContextualizingImpact2020, HennWhyDNANo2021}  But this does not mean that scientists are not responsible for the downstream social effects of their 
research, even when these effects involve deliberate misinterpretation.*[Douglas, Kitcher, Kourany, Havstad; Block and Dworkin?]*  Like other citizens, scientists are responsible for mitigating the reasonably foreseen harms that result from their actions.  And we now have decades of empirical evidence that research on the genetics of intelligence — even research that does not itself touch on race — can and will be used to promote scientific racism \cite{MeyerWrestlingSocialBehavioral2023}.  Specifically, while our findings suggest that behavior genetics has attempted to distance itself from scientific racism, behavior genetics research — and research on human genetics more generally — may be especially susceptible to appropriation by scientific racists \cite{CarlsonQuantifyingContextualizingImpact2020, OgbunugaforDNABasketballBirthday2022, *[Harden reviews]*}.

@@

\appendix

# Materials and methods

@@

![Corpus assembly. (\#fig:corpus)](img/dataset.png){width="8in" height="6.1in"}

Corpus assembly is summarized in figure \@ref(fig:corpus).  The corpus was assembled in two parts, one for *Mankind Quarterly* and the other for mainstream journals.

@@

An electronic archive of *Mankind Quarterly*, from the first issue in 1969 through 2004, is available for free on the open web at the white nationalist website *The Unz Review*.  The Python package Beautiful Soup (<https://www.crummy.com/software/BeautifulSoup/>, version not recorded) was used to retrieve every PDF available in this archive.  After identifying some gaps (missing issues) in the archive from *The Unz Review*, we manually retrieved PDFs for further articles using ProQuest.  Text was extracted using the R package `pdftools` \cite[3.0.1]{OomsPdftoolsTextExtraction2023}.  This stage of corpus assembly was conducted in Fall 2021 with the assistance of Anthony Sainez.

@@

To identify suitable mainstream journals for inclusion in the corpus, we first identified academic researchers who had been funded by the Pioneer Fund.  Pioneer is an American non-profit organization founded in 1937 to fund eugenics research and propaganda \cite{TuckerFundingScientificRacism2002}.  After the *Brown v. Board of Education* ruling in 1954, Pioneer funded various segregationist efforts across the United States, including lectures on eugenics by Stanford physicist William Shockley \cite{JacksonScienceSegregationRace2005}.  We reviewed a critical profile of Pioneer \cite{MillerPioneerFundBankrolling1994} as well as an archived page from the organization's own web site (<https://web.archive.org/web/20130103005545/http://www.pioneerfund.org/Grantees.html>), which together listed 16 researchers who had received Pioneer funds.

We then used the author search tool in Clarivate's Web of Science platform (<https://www.webofscience.com/wos/author/search>), retrieving publication lists for 14 researchers.  These searches were conducted between 2021-09-24 and 2021-10-05 by DJH.  After parsing these results, we counted how many of the 14 researchers had published in each journal.  13 journals had published 6 or more of the 14 Pioneer-funded researchers.  We excluded *Mankind Quarterly* (as already included) as well as *Science* and *Nature* (as too general) from further consideration in this side of the corpus.  3 journals published by the American Psychological Association (APA; *American Psychologist*, *Contemporary Psychology*, and *Journal of Educational Psychology*) had to be excluded due to confusion over who could give us permission to use the archives for a text mining project, with both APA and ProQuest asserting that we needed to get permission from the other entity.  *European Journal of Personality*, published by SAGE, also had to be excluded because our institutional access only went back to 1999.

The remaining 5 journals are all published by major academic publishers — Elsevier, Springer, or Cambridge University Press — and each item in the entire run of each journal has been assigned a DOI (digital object identifier) for archival purposes.  We used the Crossref API and `rcrossref` R interface to this API \cite[1.1.0.99]{ChamberlainRcrossrefClientVarious2019} to retrieve metadata for each item published from 1960-2010 in each of these journals.  These metadata included item-level license information — confirming that the text of each item could be used for text mining projects — and a URL to an electronic version of the item.  These URLs were used to retrieve a XML or PDF version of each item, except for *Personality and Individual Differences*.  This journal has published a relatively large number of non-article documents, such as book reviews and commentaries, that are not available at the URL included in the Crossref metadata.  (This is unfortunate, as it was not difficult to find highly relevant documents that we could not automatically retrieve and therefore excluded from our corpus, such as an exchange between Rushton and Flynn on the Flynn effect \cite{FlynnEvidenceRushtonGenetic1998,RushtonSecularGainsIQ1998,FlynnReplyRushtonGang1998}.)  Instead we used Elsevier's ScienceDirect API (<https://dev.elsevier.com/>) to independently search and retrieve all available items from *Personality and Individual Differences*.  This stage of corpus assembly was conducted between 2021-11 and 2022-05.

*Behavioral and Brain Sciences* typically uses a target article + commentary format; in some cases there are dozens of commentaries for a single article.  In the Crossref DOI metadata, each target article and individual commentary is given its own DOI, with no distinction between contribution types or metadata links between a commentary and the corresponding target article.  But text is only available in aggregate PDFs that bundle together the target article and all of the commentaries.  In a few cases this results in PDFs that are hundreds of pages long and might be linked from the Crossref metadata 30+ times.  In addition, the retrieved PDFs are not perfectly identical, because Cambridge UP's servers add a timestamped watermark to each page.  We attempted to contact Cambridge for assistance but did not receive a response.  We ultimately used a series of ad hoc measures to mitigate text duplication.  Approximately 100,000 documents of 1-2 pages long were excluded before PDF retrieval.  After PDF retrieval and text extraction, the timestamped watermarks were removed using a regular expression and the text was hashed using SHA 256.  Hashes were used to construct groups of duplicate documents, and a single document (whichever one happened to be first in the dataframe) was chosen from each hash group for inclusion in the corpus.  It is plausible that some duplicates made it through this process: where the watermark overlapped with the text, the regular expression may have been unable to identify and remove the watermark (a false negative result), and this difference in the watermarks (not the text) would produce different hashes.

After XML/PDF versions were retrieved, text was extracted using either the `xml2` package \cite{WickhamXml2ParseXML2023} or `pdftools`.  Figure \@ref(fig:counts) and table \@ref(tab:counts) show the number of fulltext documents included in the corpus.

![Count of documents in the corpus, by journal and year. (\#fig:counts)](img/02_count.png){width="4.76in" height="2.6in"}

Table: (\#tab:counts) Document counts, by journal, and years included in the corpus.

|journal                                |     n| start|  end|
|:--------------------------------------|-----:|-----:|----:|
|Behavior Genetics                      |  2268|  1970| 2010|
|Intelligence                           |  1237|  1977| 2010|
|Mankind Quarterly                      |  1821|  1960| 2004|
|Personality and Individual Differences |  7274|  1980| 2010|
|Psychological Reports                  | 21398|  1960| 2010|
|Behavioral and Brain Sciences          |   898|  1978| 2010|

All together, the corpus includes 34,896 documents from 6 journals.  See \@ref(fig:counts) and \@ref(tab:counts).

@@

After document retrieval and text extraction, we pre-processed the text using the `spaCy` NLP (natural language processing) Python library \cite[1.9.0]{SpaCyIndustrialstrengthNatural2018} and the R API `spacyr` \cite[1.2.1]{BenoitSpacyrWrapperSpaCy2020}.  Specifically, we applied regular expressions to remove header/footer copyright notices and hyphenation, used spaCy to annotate and extract noun phrases (eg, "the intelligence test items"), and then cleaned and standardized these phrases (eg, removing the/an/a, coverting all text to lowercase, and replacing all whitespace with underscores: "intelligence_test_items").  We then counted the occurrence of each noun phrase in the document.  The aggregated "document-term matrix" was stored in Parquet format \cite{VohraApacheParquet2016} for performance reasons, and written and read using the `arrow` package for R \cite[7.0.0]{RichardsonArrowIntegrationApache2023}.

NLP-extracted noun phrases offer a number of advantages over the more traditional unigram ("single-word") tokens.  First, noun phrase extraction removes many standard stopwords (articles, common verbs) without relying on a fixed, a priori stopword list.  Phrases can be more informative than single terms, for example, distinguishing "intelligence test" from "hypothesis test."  While simple n-gram extraction will include numerous phrases that are not especially meaningful.  Consider the sentence "Since our first analyses of feeding patterns in rats, we had been using a criterion of 40 minutes (Le Magnen & Tallon 1963; 1966)."\cite{SclafaniCorrelationCausationStudy1981}  Bigrams such as "since our" and "criterion 40" will likely be discarded in vocabulary selection, but significantly increase the computational cost of vocabulary selection.  Noun phrase extraction is therefore more efficient.

After noun phrase extraction, the corpus comprised 43,162,055 total tokens of 6,320,783 distinct phrases.

@@

We took an information-theoretic approach to vocabulary selection.\cite{HicksProductivityInterdisciplinaryImpacts2021}  Consider a game in which I draw a document from the corpus, then a single token from that document.  I tell you the term, and you have to guess which document I picked.  Intuitively, highly informative terms (in this project, noun phrase types) are distinctive, allowing you to dramatically narrow down the list of potential documents.  This "informativeness" of a term can be quantified as the KL (Kullback-Leibler) divergence from a "baseline" distribution of documents to the distribution conditional on the term.  Because the most informative terms tend to be typos and OCR errors — these are unique to a single document — we multiply the KL divergence by the logarithm of the total number of occurrences of the term across the entire corpus.  We refer to the resulting measure as $log(n) \Delta H$ or simply `ndH`.

In the current project, we found that this `ndH` approach heavily favored recurrent noun phrases in the longest documents.  Many of the documents published in *Psychological Reports* are extremely short, 1-2 page brief notices of a single study; while many of the documents published in *Brain and Behavioral Sciences* are book-length collections that include a long review article and sometimes dozens of commentaries.  Very generic noun phrases that happen to appear in the latter can occur orders of magnitude more often than highly distinctive phrases in the former, and so the $log(n)$ factor overwhelms the $\Delta H$ factor.

@@

A common rule of thumb in topic modeling is that the vocabulary should have about 10 times as many distinct terms as the number of documents in the corpus.  However, we had some concerns with computational demands here:  the resulting document-term matrix would have roughly $10 \times 33,000^2$ or 10.9 billion entries; with 10% density this would require on the order of 4 GB of memory just for a single copy of the matrix; and most of the analysis was to be conducted on the authors' laptops.  We therefore chose to work with three smaller vocabularies, $5 \times$, $1 \times$ and $\frac{1/5} \times$ the number of documents.  We refer to these as the "large," "medium," and "small" vocabulary, respectively, and they include *[counts]* distinct phrases.  We compare findings across vocabularies as a robustness check.
